---
title: "Getting Started with maplibr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with maplibr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`maplibr` is an R interface to `maplib`, a Rust RDF knowledge-graph engine
(the same engine the Python package `py_maplib` wraps). This vignette walks
through what's implemented today: building a graph, querying and updating it
with SPARQL, mapping data into RDF (both with OTTR templates and directly),
RDFS inference, and managing named graphs.

`Model` is an R6 class, so every method is also available as `m$method(...)`
-- but this vignette uses the pipe-friendly `model_*()` functions throughout
instead (`m |> model_query(...)` rather than `m$query(...)`), since both are
fully equivalent and interchangeable, and the pipe form reads better across
a multi-step pipeline. `model_*()` uses that prefix rather than bare
`read()`/`write()`/`map()`/`update()` deliberately: those names would mask
`base::write`, `purrr::map`, and the `stats::update` S3 generic if this
package were attached alongside a normal tidyverse session.

```{r setup}
library(maplibr)
```

## Creating a Model and reading triples

`Model` is the main entry point -- an in-memory RDF knowledge graph.
`model_reads()` mutates it in place and returns it invisibly, so it chains
straight off `Model$new()`:

```{r create-and-read}
m <- Model$new() |>
  model_reads('
    <http://example.org/alice> <http://www.w3.org/1999/02/22-rdf-syntax-ns#type> <http://example.org/Person> .
    <http://example.org/Person> <http://www.w3.org/2000/01/rdf-schema#subClassOf> <http://example.org/Agent> .
    <http://example.org/alice> <http://example.org/name> "Alice" .
  ', format = "ntriples")

model_size(m)
```

## Serializing back out

```{r writes}
m |> model_writes() |> cat()
```

Turtle works too:

```{r writes-turtle}
m |> model_writes(format = "turtle") |> cat()
```

`model_read()`/`model_write()` are the file-based counterparts to
`model_reads()`/`model_writes()` -- `model_read()` guesses the format from
the file extension (`.ttl`/`.nt`/`.xml`/`.rdf`) when `format` isn't given:

```{r read-write-file}
path <- tempfile(fileext = ".ttl")
m |> model_write(path, format = "turtle")

m_from_file <- Model$new() |> model_read(path)
model_size(m_from_file)
unlink(path)
```

## Querying with SPARQL

`model_query()` runs a SPARQL query and returns the result as a data.frame.

```{r query-select}
m |> model_query("SELECT ?s ?o WHERE { ?s <http://example.org/name> ?o }")
```

CONSTRUCT queries are supported too, returned as a combined
subject/predicate/object data.frame:

```{r query-construct}
m |> model_query("
  CONSTRUCT { ?s <http://example.org/hasType> ?type }
  WHERE { ?s a ?type }
")
```

`bindings` pre-binds SPARQL variables to fixed `IRI`/`Literal` values before
the query runs -- useful inside a `FILTER`, though not as a `SELECT`-ed
column itself:

```{r query-bindings}
m |> model_query(
  'SELECT * WHERE { VALUES ?name { "Alice" "Bob" } FILTER(?name = ?who) }',
  bindings = list(who = Literal("Bob"))
)
```

## Updating data with SPARQL UPDATE and INSERT

`model_update()` runs a SPARQL UPDATE (a `DELETE`/`INSERT` ... `WHERE`
modify statement) against the Model in place. The bare `INSERT DATA { ... }`
form isn't supported yet by the underlying engine -- use an empty `WHERE {}`
instead:

```{r update}
m |> model_update('
  PREFIX ex: <http://example.org/>
  INSERT { ex:alice ex:greeting "Hello" } WHERE {}
')
m |> model_query("SELECT ?o WHERE { <http://example.org/alice> <http://example.org/greeting> ?o }")
```

`model_insert()` runs a SPARQL `CONSTRUCT` query and adds its results as new
triples, optionally into a different graph than it queried:

```{r insert}
m |> model_insert(
  "CONSTRUCT { ?s <http://example.org/hasType> ?type } WHERE { ?s a ?type }",
  target_graph = "urn:graph:types"
)
model_size(m, "urn:graph:types")
```

## Mapping tabular data into RDF with OTTR templates

`model_map()` expands an OTTR template against a data.frame, adding the
resulting triples to the Model. Templates can be built programmatically...

```{r map-programmatic}
ex <- Prefix("http://example.org/", "ex")
my_value <- Variable("MyValue")
template <- Template(
  iri = suf(ex, "HasNickname"),
  parameters = list(Parameter(my_value)),
  instances = list(Triple(suf(ex, "alice"), suf(ex, "nickname"), my_value))
)

m |> model_map(template, data.frame(MyValue = c("Ally", "Al")))
m |> model_query("SELECT ?o WHERE { <http://example.org/alice> <http://example.org/nickname> ?o }")
```

...or from an stOTTR document string, matching `py_maplib`'s own syntax (see
the root project's README for the OTTR template language itself). Since
`model_add_template()` returns the registered template's IRI rather than
`m`, it's not itself chained onward -- pipe into it, but capture its result
separately from the `m` you keep mutating:

```{r map-doc-string}
doc <- '
@prefix ex:<http://example.org/>.
ex:HasFriend [?Friend] :: {
  ottr:Triple(ex:alice, ex:knows, ?Friend)
} .
'
m |> model_add_template(doc)
m |> model_map("http://example.org/HasFriend", data.frame(Friend = "http://example.org/bob"))
m |> model_query("SELECT ?o WHERE { <http://example.org/alice> <http://example.org/knows> ?o }")
```

## Mapping data directly to triples, without a template

For data that doesn't need OTTR's templating power, several `model_map_*()`
functions add triples directly instead. `model_map_df()` turns a
data.frame's columns into predicates, one row per subject:

```{r map-df}
readings <- data.frame(city = c("Oslo", "Bergen"), temp_c = c(14, 16))
m_direct <- Model$new() |> model_map_df(readings)
m_direct |> model_query("SELECT ?p ?o WHERE { ?s ?p ?o }")
```

`model_map_triples()` maps subject/predicate/object columns directly (or
just subject/object, with a constant `predicate`):

```{r map-triples}
m_triples <- Model$new() |> model_map_triples(
  data.frame(subject = "http://example.org/alice", object = "http://example.org/bob"),
  predicate = "http://example.org/knows"
)
m_triples |> model_query("SELECT ?s ?p ?o WHERE { ?s ?p ?o }")
```

`model_map_json_string()`/`model_map_xml_string()` (and their file-based
`model_map_json()`/`model_map_xml()` counterparts) apply a fixed convention
(Facade-X) to a whole JSON or XML document instead:

```{r map-json}
m_json <- Model$new() |> model_map_json_string('{"name": "Alice", "age": 30}')
m_json |> model_query("SELECT ?p ?o WHERE { ?s ?p ?o }")
```

## RDFS inference

`alice` is a `Person`, and `Person` is a subclass of `Agent` -- RDFS
inference should let us derive that `alice` is also an `Agent`.
`model_infer_rdfs()` returns the number of newly-inferred triples, not `m`,
so this one's called directly rather than piped further:

```{r infer}
n <- m |> model_infer_rdfs()
n
```

Inferred triples are *transient* -- kept separately from the main graph, and
only visible to `model_query()` when asked for explicitly (`model_writes()`
doesn't currently support this at all, a core-engine limitation, not an
R-wrapper one -- see the "Upstream issues" section of the package's
development notes):

```{r infer-transient}
without <- m |> model_query("SELECT * WHERE { ?s ?p ?o }")
with_transient <- m |> model_query("SELECT * WHERE { ?s ?p ?o }", include_transient = TRUE)
nrow(without)
nrow(with_transient)
```

## Named graphs

Triples can be routed into named graphs, copied between Models, and split
back out.

```{r graphs}
m |> model_reads(
  '<http://example.org/bob> <http://example.org/name> "Bob" .',
  format = "ntriples",
  graph = "urn:graph:extra"
)
model_size(m, "urn:graph:extra")
model_size(m) # default graph, unaffected

m2 <- Model$new() |> model_add_graph(m)
model_size(m2)

sprout <- m |> model_detach_graph(graph = "urn:graph:extra")
model_size(sprout)
model_size(m, "urn:graph:extra") # emptied by the detach
```

## Prefixes and indexing

Both mutate in place, so a two-step setup pipeline chains naturally:

```{r prefixes-index}
m |>
  model_add_prefixes(list(ex = "http://example.org/")) |>
  model_create_index()
```

## Visualizing a graph

`maplibr` has no graph-visualization function of its own, but a query result
is a plain data.frame, so any R graph package works directly on it.

```{r graph-plot, fig.width=6, fig.height=4.5}
if (requireNamespace("igraph", quietly = TRUE)) {
  triples <- m |> model_query("SELECT ?s ?p ?o WHERE { ?s ?p ?o }")
  short_label <- function(iri) sub(".*[/#]", "", iri)

  edges <- data.frame(from = short_label(triples$s), to = short_label(triples$o))
  g <- igraph::graph_from_data_frame(edges, directed = TRUE)
  igraph::E(g)$label <- short_label(triples$p)

  plot(
    g,
    edge.label = igraph::E(g)$label,
    vertex.size = 32,
    vertex.color = "lightblue",
    vertex.label.cex = 0.9,
    edge.arrow.size = 0.5,
    layout = igraph::layout_with_fr(g)
  )
}
```
